---
title: "metaboCombineR - A tool for combining untargeted multi-batch
LC-MS experiments"
author: "Frantisek Malinka"
output:
BiocStyle::html_document:
toc_float: true
vignette: >
%\VignetteIndexEntry{metaboCombineR}
%\VignetteEngine{knitr::rmarkdown}
%\VignetteEncoding{UTF-8}
%\VignetteDepends{metaboCombineR, BiocStyle}
%\VignettePackage{metaboCombineR}
%\VignetteKeywords{LC/MS, elution order, peak}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r init, message = FALSE, echo = FALSE, results = "hide" }
## Silently loading all packages
library(BiocStyle)
library(metaboCombineR)

```

**Package**: `r Biocpkg("metaboCombineR")`<br />
**Authors**: Frantisek Malinka<br />
**Modified**: `r file.info("metaboCombineR")$mtime`<br />
**Compiled**: `r date()`

# Introduction

This document describes R package `metaboCombineR` and demonstrates its
suitability and usage on an authentic untargeted LC-MS dataset. metaboCombineR
is written in C++ focusing on large-scale untargeted LC-MS experiments.
The package allowes to preprocessed each batch of samples separately using
an arbitrary preprocessing program, such as XCMS. The main pros is a possibility
to handle and observed results of experiments during the time and program
parameter tuning should be easier since within-batch variablity is smaller
than beetween-batch variablity. The package suppose only basic R skills.

# Data format

The input data is simply a list of numerical 2-dimensional matrices
(`data.frame` structure in R) where rows correspond to LC-MS features and
columns correspond to samples in the given batch. Therefore, each matrix
is a result of previously preprocessed samples from a batch. Note that both
dimensions must be named. Name of columns, samples, is arbitrary. However, name
of each row, feature,  consist of m/z value and retention time. The m/z value is
prefixed by `M` and the retention time is prefixed by `T`. For example, a
feature with m/z value equal 57.08131 and retention time equal 1428.18786 is
written as `M57.08131T1428.18786`.

## Loading example datasets 
In this package, we provide four example datasets for demonstrating a
functionality of implemented method. To load these datasets into your R
workspace, call `data` function.
```{r example datasets}
data(metaboExp1)
data(metaboExp2)
data(metaboExp3)
data(metaboExp4)
```

Now, all datasets are ready to use. To confirm that datasets are in the
required format, check the structure of input `data.frame`s. Rows nad columns
must be named and features must be in the correct format.
```{r datmatrix example, echo=TRUE}
head(metaboExp1)
```

To verify it, we provide three functions: `getMZs`, `getRTs`,
`sortDataFrameByRt`, and `sortDataFrameByMz`. Obviously from the names,
`getMZs` function returns a vector of m/z values, `getRTs` function returns
a vector of values corresponding to retention time.

```{r getRTs and getMZs}
mzs <- getMZs(metaboExp1)
rts <- getRTs(metaboExp1)
head(mzs)
head(rts)
```

`sortDataFrameByRt` function returns a data.frame where rows are sorted
by the corresponding rt values.
```{r sort data.frame by rt}
sortedRTdf <- sortDataFrameByRt(metaboExp1)
head(sortedRTdf)
```

`sortDataFrameByMz` function returns a data.frame where rows are sorted
by the corresponding mz/s values.
```{r sort data.frame by mz}
sortedMZdf <- sortDataFrameByMz(metaboExp1)
head(sortedMZdf)
```

## Run metaboCombineR
Finally, to run the method called `runMetaboCombiner` which requires 3
arguments, i.e. `listExperimens` must be a list of data.frames representing
experiments, `mzprecision`  expressed number of mz digits to be considered
(default is 3), `windowsize` set up length of subsequences (default is 5).

```{r runall, eval=FALSE}
runMetaboCombiner(list(metaboExp1, metaboExp2, metaboExp3, metaboExp4),
windowsize = 5)
```
